Pseudogap and Mott criticality: Stretching Kondo Screening to Breaking Point

### 2025-12-10 submissions/TilingUploadNJPResponse/RefereeResponse.tex

\documentclass[10pt]{revtex4-2}
\usepackage{amsmath,amssymb,braket,newtxtext}
\usepackage[hidelinks]{hyperref}
\usepackage{color}
\usepackage{xcolor}
% \usepackage[margin=1.5cm,bottom=1.5cm]{geometry}
\newcommand{\shat}{\hat{\mathbf{s}}}
\newcommand{\response}[1]{{\bf Response. }{\color{blue}{#1}}}

\begin{document}

\title{RESPONSE TO REFEREE COMMENTS FOR NJP-119431}
\author{Abhirup Mukherjee}
\author{S. R. Hassan}
\author{Anamitra Mukherjee}
\author{N. S. Vidhyadhiraja}
\author{A. Taraphder}
\author{Siddhartha Lal}
\maketitle
\vspace{-20pt}
\hrulefill
\vspace{10pt}

We thank the Editorial Board Member and all referees for their careful reading of our manuscript and for their constructive comments and questions. We are encouraged by the overall positive reception of our work, including the acknowledgement of its relevance, technical soundness, and suitability for publication in New Journal of Physics.

Below, we present our responses (in blue) to the points raised by the referees.

\section*{Response to Referee 1}
The authors proposed the possibility that the pseudogap phase is a distinct quantum phase characterized by the so-called Mott metal. I recommend sending the manuscript to reviewers.

\section*{Response to Referee 2}
The impurity-to-lattice tiling procedure and unitary renormalization group (URG) methods, while carefully implemented, appear to extend prior impurity-based constructions rather than introduce a fundamentally new paradigm. The claimed "controlled route to Mott criticality" should be more clearly distinguished from existing dynamical mean-field theory (DMFT) and functional renormalization group (fRG) approaches, particularly in terms of predictive power or access to uncharted regimes.

\response{}

The identification of the pseudogap as a "Mott metal" with deconfined holon-doublon excitations is intriguing but lacks direct experimental or numerical validation beyond the model itself. Comparisons with established benchmarks from quantum Monte Carlo, ARPES, or optical conductivity experiments on cuprates would help ground these claims and clarify the added value beyond prior theoretical proposals.

\response{hi}

The restriction to a single-band extended Hubbard model on a square lattice limits the generality of the conclusions. Demonstrating the framework's applicability to at least one additional model or geometry would strengthen its broad relevance to correlated systems.

\response{}

Critical experimental realities such as disorder, doping inhomogeneity, or phonon coupling are not addressed. A discussion of how such effects could be incorporated or might alter the conclusions is necessary to assess the experimental relevance of the proposed mechanism.

\response{hi}

The connection to the Hatsugai-Kohmoto model at criticality, while mathematically interesting, risks being viewed as a special case. The authors should clarify whether this connection holds for generic models or relies on fine-tuned symmetries or fillings.

\response{hi}

\section*{Response to Referee 3}
In the present manuscript the authors have studied the interplay between the spin exchange and local Coulomb interaction in a two-dimensional model by employing a unitary renormalization group (URG). This work can be considered as a continuation of the author's previous published works. While I found that the URG is acceptable for treating electron correlation problems, the principal model studied in this manuscript (Eq. (15)) looks like an artefact, because the spin exchange  (the second term in Eq. (15)) is actually can be generated by the hopping and local interaction terms. In general, the spin exchange is roughly proportional to $t^2/U$ (t, U here correspond to the ones with tilde symbol in Eq. (15)). Therefore, the studied model already double-counts the spin exchange. In addition, it is not clear what real materials or systems can be described by the studied model. The authors have to clarify these points before the manuscript can be acceptable for publication.

\section*{Response to Referee 4}
The manuscript addresses metal-to-insulator transition in strongly correlated lattice models applicable to materials such as layered copper oxides, with particular focus on intermediate pseudogap (PG) state. First, a lattice impurity model is formulated, and subsequently mapped to a lattice t-J-U model via tiling procedure. The Authors subsequently map the Green's functions from the auxiliary impurity model to the target t-J-U model, and analyze its dynamical phase diagram as a function of microscopic parameters. The employed methodology is based on solving normalization group equations for the auxiliary model and translating the results into the of target lattice model language via derived mapping. The Authors link the PG state to the to the breakdown of the Kondo screening within the underlying auxiliary model, and provide a comprehensive theoretical analysis of thereof.

I have looked through it with interest, and I have several comments and queries:

1. The Authors consider the model with nearest-neighbor hopping (t') integrals only, which is also reflected in the target lattice model. It is well known that t' is highly relevant to the physics of systems exhibiting pseudogap (such as high-Tc cuprates). Is it possible to incorporate those terms within the present method and is there any qualitative change expected?
2. The Hamiltonian term $H_coup$ [Eq. (3)] incorporates both hybridization and Kondo-type exchange. Within an impurity model, the former Kondo term is typically derived by canonical perturbation expansion from he model containing V-term, so (technically) including both may be regarded as double-counting. Can, for instance, either V or J term be neglected in the analysis? What would be the consequences for PG state?
3. Technical remarks: In Eq. (3), J should be probably a function of Z to reproduce k-dependen form of J in Eq. (5). Apparently in Eq. (A1) of the first Appendix this is the case. Also, it is not immediately clear to me why the impurity creation and annihilation operators disappear from the impurity model (1) following tiling procedure in Sec. III, resulting in single-band Hamiltonian (15). This could be explained in grater detail in the manuscript.
4. The lattice model of Eq. (15), referred to by the Authors as extended-Hubbard, has been extensively studied in literature in the context of high-Tc superconductors as the t-J-U model. See, e.g., https://doi.org/10.1103/PhysRevB.95.024506 or a review https://doi.org/10.1016/j.physrep.2022.02.003 and references therein. Comparison with previous work on t-J-U model would benefit the manuscript.
5. Whereas the Authors work is clearly relevant to specific materials, such and layered copper oxides, nickelates, and possibly several other systems, the discussion of specific experimental consequences is limited. The manuscript would benefit from more direct and detailed relation to specific materials in the introduction and/or summary section.

The manuscript is well written and organized. The methodology appears sound and the mapping of the auxiliary to physical model is of interest. I believe that the manuscript could be suitable for publication in NJP, following clarification of the above points.

\end{document}

### 2025-12-10 submissions/Earlier Versions of Manuscript/manuscript2.tex
started creating referee response

% \documentclass[%
%  reprint,
% superscriptaddress,
% groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
% amsmath,amssymb,
% aps,
% prl,superscriptaddress
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
% pdflatex,
% sn-nature,
% ]{sn-jnl}
\documentclass[pdflatex,sn-mathphys-num]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{braket}
\usepackage{color}

\graphicspath{{./figures/}}

@@

\title{Pseudogap Mott criticality via a long-range multipartite entangled strongly coupled non-Fermi liquid}

%\title{Pseudogap and Mott criticality: Stretching Kondo Screening to Breaking Point}

@@

\author[1]{Abhirup Mukherjee}
\affil[1]{
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}
\author[2]{S. R. Hassan}
\affil[2]{The Institute of Mathematical Sciences, C.I.T. Campus, Chennai 600 113, India}
\author[3]{Anamitra Mukherjee}
\affil[3]{School of Physical Sciences, National Institute of Science, Education and Research, HBNI, Jatni 752050, India}
\author[4]{N. S. Vidhyadhiraja}
\affil[4]{Theoretical Sciences Unit, Jawaharlal Nehru Center for Advanced Scientific Research, Jakkur, Bengaluru 560064, India}
\author[5]{A. Taraphder}
\affil[5]{Department of Physics, Indian Institute of Technology Kharagpur, Kharagpur 721302, India}
%
\author*[1]{Siddhartha Lal}

\date{\today}
\abstract{
We propose a new theoretical framework for strongly correlated electron systems 
%We present an auxiliary model approach 
that illustrates the centrality of the pseudogap phenomenon in the Mott transition of the extended Hubbard model on a half-filled square lattice. Our approach is based on %Vacancy--Kondo--Heisenberg (VKH) model and a Quantum Tiling Reconstruction (QTR) scheme. 
%a real-space Kondo-defect unit ---comprising a localized 
a spin-1/2 defect at a lattice vacancy that is Kondo exchange-coupled to neighbouring lattice sites, and which is tiled across a square lattice to yield a translationally invariant model. The pseudogap is emergent from a Fermi liquid via the systematic unravelling of Kondo screening frustrated by charge fluctuations in the conduction bath of the underlying lattice-embedded Kondo model. We demonstrate that the pseudogap is a strongly interacting phase of quantum matter intimately tied to the existence of a critical Fermi surface, with non-Fermi excitations that have long-ranged and multipartite entanglement. The Mott criticality of the non-Fermi liquid is signalled by the emergence of a singular self-energy at the Fermi energy. Our analysis points to this novel phase of gapless quantum matter being stabilised by changes in the topological properties of the Fermi surface that arise from encroaching surfaces of zeros of the single-particle Greens function.
%passage from a Fermi liquid to a Mott insulator involves two stages: the first involves a systematic unravelling of Kondo screening, frustrated by charge fluctuations in the conduction bath of the underlying lattice-embedded quantum impurity model. 
%This destabilises the Fermi liquid, culminating in the disconnection of the Fermi surface with the appearance of gaps at the antinodes. In the second stage, an emergent pseudogap phase arises from an effective two-channel Kondo problem in the underlying impurity model. The resulting Fermi arcs possess non-Fermi liquid excitations proximate to a critical Fermi surface. Upon approaching the Mott transition, the nature of this exotic metal evolves with shrinking of the Fermi arcs towards a singular nodal non-Fermi liquid. This accompanies the onset of long-range multipartite entanglement and spin-flip correlations in real-space, revealing the Mott transition on the square lattice to be beyond the paradigm of local quantum criticality.
}

@@

\section{Introduction.}
The origin of the pseudogap and strange metal phases of the cuprates continues to be hotly debated in the context of high-temperature superconductivity~\cite{keimer2015quantum}. The anomalous pseudogap (PG) phase, observed in the cuprates, exhibits nodal-antinodal dichotomy with spectral weight concentrated on Fermi arcs around the nodal regions~\cite{loeserKapitulnik1996,Norman1998,Hashimoto2014}. While several calculations have observed a PG in the two dimensional Hubbard model ~\cite{KyungKotliar2006,MacridinAzevedo2006,WuFerrero2018,anirbanmott2,HilleAndergassen2020}, there is no general consensus regarding various aspects of the PG phase, including its relation to the proximate Mott insulating and superconducting phases~\cite{FradkinRevModPhys2015,anirbanmott2,Kitatani2023,Sorella2023}, its evolution from weak- to strong-coupling ~\cite{HuangDevereaux2018,Fedor2022,SimkovicFerrero2024}, and whether the nodal-antinodal dichotomy is intrinsic to the model~\cite{Hashimoto2014,Schafer2021}. Importantly, the connection between finite-temperature crossovers and zero-temperature ground states observed in various analyses remains unclear~\cite{White1998,Ido2018,ProustTaillefer2019,Ponsioen2019,Shengtao2021,XuZhang2022}.

Viewed as a quantum mechanical phenomenon, the PG involves the emergence of disconnected gapped regions of the Fermi surface (known as Luttinger surfaces of zeros of the single-particle Greens function ~\cite{dzyaloshinskii2003some,phillips2006mottness}) from a Fermi liquid with a connected Fermi surface. The growth and eventual connection of these Luttinger surfaces then leads to the transition into the Mott insulator. Which physical principles guide these events and what do they teach us about the criticality that obtains Mott insulation? Further, can the excitations of the gapless Fermi arcs be adiabatically connected to those of a non-interacting electron gas in the presence of the Luttinger surfaces? If not, what kind of scale-invariant quantum matter do they characterise? Which features of the PG elucidate the breakdown of screening in the Fermi liquid and lead instead to a singular self-energy in the Mott insulator?

We present a new auxiliary model approach 
to the strong coupling physics of particle-hole symmetric 2D Hubbard-Heisenberg model at zero temperature, revealing the origin and nature of its PG phase and the subsequent Mott transition. The first step 
requires identifying an appropriate lattice-embedded quantum impurity model, and a scaling analysis of its low-energy phases. Subsequently, a lattice model is  constructed by applying many-body lattice translation operators~\cite{stoyanova} on the impurity model (henceforth referred as \textit{tiling}), obtaining thereby various quantities of the former from the latter. Translation invariance is thus restored differently within our approach from the dynamical mean-field approximation~\cite{georges1996} and its cluster variants~\cite{Hettler2000DCA, Kotliar2001Cellular, Maier2005, Sakai2023}.

Combining the scaling analysis of the impurity model with tiling procedure provides momentum-space resolution crucial to unveiling the PG phenomenon, and clarifies the role of the Kondo breakdown process at the heart of Mott transition~\cite{fabrizio2017kondophysicsmotttransition,RADEMAKER2025}. 
Emerging from a systematic unraveling of Kondo screening, the PG phase is separated from the Fermi liquid (FL) and Mott insulating phases by correlation-driven Lifshitz transitions of the Fermi surface (FS)~\cite{sakai2009evolution}. Passage through the PG involves dynamical transfer of spectral weight between gapped anti-nodal Luttinger surfaces (zeros of the single-particle Green's function) and gapless nodal arcs with non-FL excitations. We also observe the concomitant onset of long-range real-space entanglement and spin-flip correlations close to the Mott transition, rendering it beyond local quantum criticality~\cite{Si2001}. Our findings serve as a demonstration of the PG as a strongly interacting long-range multipartite entangled phase of quantum matter linked to the Mott criticality of the 2D extended Hubbard model~\cite{Phillips2013Unparticles}. This follows a recent proposal~\cite{lanave2025} that the emergence of Luttinger surfaces lead to the spontaneous violation of a $\mathbb{Z}_{2}$ symmetry associated with the Fermi surface~\cite{Anderson2001,Huang2022}, and thence to the existence of non-local propagating degrees of freedom.

\textcolor{red}{ AM: Some comments expanding on for identifying a underlying principle: The encroaching of the Fermi surface by the creeping Luttinger surface increases with electronic interaction between the bath electrons (W), which frustrates the Kondo screening and is controlled by (W/J). This phenomenon leads to a transition to a non-Fermi liquid state, where the quasiparticle excitations lose coherence and are characterized by disconnected Fermi surfaces and a pseudo-gapped density of states. The eventual demise of the Fermi surface and its replacement by a connected Luttinger surface finally leads to the Mott phase. }

\textcolor{red}{We demonstrate that a nodal non-Fermi liquid (at the verge of the Mott transition, when the Fermi surface has shrunk to a point) is described by the exactly solvable Hatsugai-Kohmoto (HK) model. The (HK) model exhibits a specific type of symmetry breaking, which is crucial for stabilizing non-Fermi liquids and Mott physics. This involves breaking a local-in-momentum space $Z_2$ symmetry, which serves as an organizing principle for the Mott transition. This $Z_2$ symmetry [Ansderon, Haldane,  Phillips] is local-in-momentum space and non-local in real space (exhibiting long-range multi-partite entanglement (it would be nice if we could demonstrate this) ). When this symmetry is disrupted, the symmetry between particle addition and removal around the chemical potential is also lost. The resulting non-Fermi liquid is characterized by a self-energy scaling as $\omega^{-2}$ with energy in the HK model. We show that the same self-energy characterizes the entire pseudo-gapped regime of the extended Hubbard-Heisenberg model. We also demonstrate the $Z_2$ breaking for the URG-derived Hamiltonian in the pseudo-gapped phase, concretely linking the origin of the non-Fermi liquid behavior to the loss of particle-hole symmetry.  Our result reveals a new adiabatic continuity connecting the Mott insulator phase (described by the HK model by the emergence of a fully connected Luttinger surface) to a family of pseudo-gapped non-Fermi liquid states, in analogy to adiabatic continuity of the noninteracting fermions to the Fermi liquid phase.}

\section{A lattice-embedded impurity model.}

@@

\section{Reconstructing the lattice model.}

@@

    \includegraphics[width=0.325\linewidth]{phaseDiagram-77-0.001.pdf}
    \includegraphics[width=0.325\linewidth]{SF.pdf}
    \includegraphics[width=0.325\linewidth]{kspaceDOS-77.pdf}

@@

\section{Pseudogap Formation via Kondo Breakdown.}
A detailed picture of Kondo breakdown in the large $U$ phase of $H_\text{aux}$ is obtained from a scaling analysis of the ${\bf k}$-resolved Kondo coupling \(J^{(j)}_{{\bf k}_1, {\bf k}_2}\) using the 
unitary renormalisation group (URG) method \cite{anirbanurg1} (see \cite{suppmat} for details)
\begin{equation}\begin{aligned}\label{KondoRGequation}
	\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2}~,
where \(\varepsilon_j\) is the energy of the shell being decoupled at the \(j^\text{th}\) step, the sum is over all occupied momentum states \({\bf q}\) 
%in the particle sector (PS {\textcolor{green}{are we using the abbreviation PS anywhere else?}}) 
of the energy shell \(\varepsilon_j\), 
%(i.e., all states occupied at \(T=0\) and in the absence of any quantum fluctuations), 
and  \({\bf \bar q} = {\bf q} + {\boldsymbol \pi}\) is the particle-hole transformed state associated with ${\bf q}$. The bath interaction coupling $W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}$ is found to be marginal under these transformations. While the complexity of the RG equation for $J_{{\bf k}_1, {\bf k}_2}$ (eq.\eqref{KondoRGequation}) is clarified through a detailed numerical evaluation whose results are discussed below, its structure permits the broad conclusion that the frustration of Kondo screening due to charge fluctuations (for attractive bath interactions $W<0$) lead to the Mott metal-insulator transition~\cite{Mukherjee_2023}.

Upon tuning the ratio of the bath and Kondo interactions ($W/J$) from zero to negative values (see phase diagram in Fig.~\ref{spinCorr}(left)), the following phases emerge in the impurity model from the competition between $J$ and $W$ in eq.~\eqref{KondoRGequation}: (i) for $W/J<(W/J)_{\text{PG}}$, an LFL phase (where the entire FS participates in Kondo screening), (ii) for $\frac{W}{J} \in [(\frac{W}{J})_{\text{PG}}, (\frac{W}{J})_c]$, a local PG phase where disconnected parts of the FS around the node participate in Kondo screening, and (iii) a local moment phase for $\frac{W}{J} > (\frac{W}{J})_c$, where the impurity remains unscreened at low-energies. These can be visualised from spin correlations $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$; see Fig.~\ref{spinCorr}(center) for $\chi_s(d,{\bf k})$ in the PG. The values $(W/J)_{\text{PG}}$ and $(W/J)_c$ are therefore the entry into and exit from the PG phase. Mapping on to the lattice model via tiling, we observe that the $T=0$ Mott transition of the 2D Hubbard-Heisenberg model proceeds from FL to Mott insulator through an intervening PG phase having a partially gapped FS (i.e., a vanishing electronic density of states around the antinodal regions, Fig.~\ref{spinCorr}(right)). We now outline the mechanism for the formation of the PG.

>>>>

\documentclass[%
 reprint,
superscriptaddress,
groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
prl,superscriptaddress
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4-2}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{braket}
\usepackage{color}
\graphicspath{{./figures/}}
\bibliographystyle{apsrev4-2}

@@

%\title{Long range entangled Fermi arcs at a Pseudogapped Mott transition}

\title{Pseudogap and Mott criticality: Stretching Kondo Screening to Breaking Point}

@@

\author{Abhirup Mukherjee}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}
\author{S. R. Hassan}
\affiliation{The Institute of Mathematical Sciences, C.I.T. Campus, Chennai 600 113, India}
\author{Anamitra Mukherjee}
\affiliation{School of Physical Sciences, National Institute of Science, Education and Research, HBNI, Jatni 752050, India}
\author{N. S. Vidhyadhiraja}
\affiliation{Theoretical Sciences Unit, Jawaharlal Nehru Center for Advanced Scientific Research, Jakkur, Bengaluru 560064, India}
\author{A. Taraphder}
\affiliation{Department of Physics, Indian Institute of Technology Kharagpur, Kharagpur 721302, India}
%
\author{Siddhartha Lal}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}

\date{\today}
\begin{abstract}
We present an auxiliary model approach that illustrates the centrality of the pseudogap phenomenon in the Mott transition of the Hubbard-Heisenberg model on a half-filled square lattice. The passage from a Fermi liquid to a Mott insulator involves two stages: the first involves a systematic unravelling of Kondo screening, frustrated by charge fluctuations in the conduction bath of the underlying lattice-embedded quantum impurity model. This destabilises the Fermi liquid, culminating in the disconnection of the Fermi surface with the appearance of gaps at the antinodes. In the second stage, an emergent pseudogap phase arises from an effective two-channel Kondo problem in the underlying impurity model. The resulting Fermi arcs possess non-Fermi liquid excitations proximate to a critical Fermi surface. Upon approaching the Mott transition, the nature of this exotic metal evolves with shrinking of the Fermi arcs towards a singular nodal non-Fermi liquid. This accompanies the onset of long-range real-space entanglement and spin-flip correlations, revealing the Mott transition on the square lattice to be beyond the paradigm of local quantum criticality.
\end{abstract}

@@

\par\noindent\textit{Introduction.}
The origin of the pseudogap and strange metal phases of the cuprates continues to be hotly debated in the context of high-temperature superconductivity~\cite{keimer2015quantum}. The anomalous pseudogap (PG) phase, observed in the cuprates, exhibits nodal-antinodal dichotomy with spectral weight concentrated on Fermi arcs around the nodal regions~\cite{loeserKapitulnik1996,Norman1998,Hashimoto2014}. While several calculations have observed a PG in the two dimensional Hubbard model ~\cite{KyungKotliar2006,MacridinAzevedo2006,WuFerrero2018,anirbanmott2,HilleAndergassen2020}, there is no general consensus regarding various aspects of the PG phase, including its relation to the proximate Mott insulating and superconducting phases~\cite{FradkinRevModPhys2015,anirbanmott2,Kitatani2023,Sorella2023}, its evolution from weak- to strong-coupling ~\cite{HuangDevereaux2018,Fedor2022,SimkovicFerrero2024}, and whether the nodal-antinodal dichotomy is intrinsic to the model~\cite{Hashimoto2014,Schafer2021}. Importantly, the connection between finite-temperature crossovers and zero-temperature ground states observed in various analyses remains unclear~\cite{White1998,Ido2018,ProustTaillefer2019,Ponsioen2019,Shengtao2021,XuZhang2022}.

\textcolor{blue}{Viewed as a quantum mechanical phenomenon, the PG involves the emergence of disconnected gapped regions of the Fermi surface (known as Luttinger surfaces of zeros of the single-particle Greens function (CITE references)) from a Fermi liquid with a connected Fermi surface. The growth and eventual connection of these Luttinger surfaces then leads to the transition into the Mott insulator. Which physical principles guide these events and what do they teach us about the criticality that obtains Mott insulation? Further, can the excitations of the gapless Fermi arcs be adiabatically connected to those of a non-interacting electron gas in the presence of the Luttinger surfaces? If not, what kind of scale-invariant quantum matter do they characterise? Which features of the PG elucidate the breakdown of screening in the Fermi liquid and lead instead to a singular self-energy in the Mott insulator?}

We present a new auxiliary model approach 
to the strong coupling physics of particle-hole symmetric 2D Hubbard-Heisenberg model at zero temperature, revealing the origin and nature of its PG phase and the subsequent Mott transition. The first step 
requires identifying an appropriate lattice-embedded quantum impurity model, and a scaling analysis of its low-energy phases. Subsequently, a lattice model is  constructed by applying many-body lattice translation operators~\cite{stoyanova} on the impurity model (henceforth referred as \textit{tiling}), obtaining thereby various quantities of the former from the latter. Translation invariance is thus restored differently within our approach from the dynamical mean-field approximation~\cite{georges1996} and its cluster variants~\cite{Hettler2000DCA,Kotliar2001Cellular,Maier2005,Sakai2023}.

Combining the scaling analysis of the impurity model with tiling procedure provides momentum-space resolution crucial to unveiling the PG phenomenon, and clarifies the role of the Kondo breakdown process at the heart of Mott transition~\cite{fabrizio2017kondophysicsmotttransition,RADEMAKER2025}. 
Emerging from a systematic unraveling of Kondo screening, the PG phase is separated from the Fermi liquid (FL) and Mott insulating phases by correlation-driven Lifshitz transitions of the Fermi surface (FS)~\cite{sakai2009evolution}. Passage through the PG involves dynamical transfer of spectral weight between gapped anti-nodal Luttinger surfaces (zeros of the single-particle Green's function) and gapless nodal arcs with non-FL excitations. We also observe the \textcolor{blue}{concomitant} onset of long-range real-space entanglement and spin-flip correlations close to the Mott transition, rendering it beyond local quantum criticality~\cite{Si2001}. \textcolor{blue}{Our findings serve as a demonstration of a recent proposal (CITE La Nave etal. arXiv:2506.04342) that disconnected Luttinger surfaces leads to the spontaneous violation of a $\mathbb{Z}_{2}$ symmetry associated with the Fermi surface (CITE HALDANE-ANDERSON, Recent work by Philips etal), and thence to the existence of non-local propagating degrees of freedom.}

\par\noindent\textit{A lattice-embedded impurity model.}

@@

\par\noindent\textit{Reconstructing the lattice model.}

@@

    \includegraphics[width=0.31\linewidth, height=2.45cm]{phaseDiagram.pdf}
    \includegraphics[width=0.67\linewidth]{SF-DOS.pdf}

@@

\par\noindent\textit{Pseudogap Formation via Kondo Breakdown.}
A detailed picture of Kondo breakdown in the large $U$ phase of $H_\text{aux}$ is obtained from a scaling analysis of the ${\bf k}$-resolved Kondo coupling \(J^{(j)}_{{\bf k}_1, {\bf k}_2}\) using the 
unitary renormalisation group (URG) method \cite{anirbanurg1} (see \cite{suppmat} for details)
\begin{equation}\begin{aligned}\label{KondoRGequation}
	\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2}~,
\end{aligned}\end{equation}
where \(\varepsilon_j\) is the energy of the shell being decoupled at the \(j^\text{th}\) step, the sum is over all occupied momentum states \({\bf q}\) 
%in the particle sector (PS {\textcolor{green}{are we using the abbreviation PS anywhere else?}}) 
of the energy shell \(\varepsilon_j\), 
%(i.e., all states occupied at \(T=0\) and in the absence of any quantum fluctuations), 
and  \({\bf \bar q} = {\bf q} + {\boldsymbol \pi}\) is the particle-hole transformed state associated with ${\bf q}$. The bath interaction coupling $W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}$ is found to be marginal under these transformations. While the complexity of the RG equation for $J_{{\bf k}_1, {\bf k}_2}$ (eq.\eqref{KondoRGequation}) is clarified through a detailed numerical evaluation whose results are discussed below, its structure permits the broad conclusion that the frustration of Kondo screening due to charge fluctuations (for attractive bath interactions $W<0$) lead to the Mott metal-insulator transition~\cite{Mukherjee_2023}.

Upon tuning the ratio of the bath and Kondo interactions ($W/J$) from zero to negative values (see phase diagram in Fig.\ref{spinCorr}(left)), the following phases emerge in the impurity model from the competition between $J$ and $W$ in eq.~\eqref{KondoRGequation}: (i) for $W/J<(W/J)_{\text{PG}}$, an LFL phase (where the entire FS participates in Kondo screening), (ii) for $\frac{W}{J} \in [(\frac{W}{J})_{\text{PG}}, (\frac{W}{J})_c]$, a local PG phase where disconnected parts of the FS around the node participate in Kondo screening, and (iii) a local moment phase for $\frac{W}{J} > (\frac{W}{J})_c$, where the impurity remains unscreened at low-energies. These can be visualised from spin correlations $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$; see Fig.\ref{spinCorr}(center) for $\chi_s(d,{\bf k})$ in the PG. The values $(W/J)_{\text{PG}}$ and $(W/J)_c$ are therefore the entry into and exit from the PG phase. Mapping on to the lattice model via tiling, we observe that the $T=0$ Mott transition of the 2D Hubbard-Heisenberg model proceeds from FL to Mott insulator through an intervening PG phase having a partially gapped FS (i.e., a vanishing electronic density of states around the antinodal regions, Fig.\ref{spinCorr}(right)). We now outline the mechanism for the formation of the PG.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\linewidth]{zerosFlow.pdf}
    \caption{Left, Right and Center: Decoupling of $J_{{\bf k}, {\bf k'}}$ (positive, negative and zeros shown in yellow, purple and green respectively), with ${\bf k}$ (grey circle) corresponding to the node, antinode and a point mid-way between them on the top right arm of the FS respectively. This is seen via the appearance of zeros (green patches) of $J_{{\bf k}, {\bf k'}}$ for ${\bf k'}$ initially on the nodal regions of adjacent arms, and their subsequent progression towards the antinodes. The decoupling 
    ends with the onset of the pseudogap.
    }
    \label{rgProgression}
\end{figure}
%\newline 
\par\noindent\textit{Unraveling of Kondo screening.} 
As shown in Fig.\ref{rgProgression}, the anisotropy in the ${\bf k}$-space of Kondo breakdown can be visualized in terms of zeros of $J_{{\bf k}_N, {\bf k}}$, involving spin-flip scattering between the node ${\bf k}_N = (\pi/2, \pi/2)$  and a general wavevector ${\bf k}$. 
For any $W/J$, the $\mathcal{C}_{4}$ lattice symmetry dictates that $J_{{\bf k}_N, {\bf k}}$ vanishes if ${\bf k}$ belongs to any of the antinodes or adjacent nodes. Tuning $W/J$ towards $(W/J)_{\text{PG}}$ leads to an unraveling of the Kondo screening: $J_{{\bf k}_N, {\bf k}}$ for ${\bf k}$ close to the adjacent nodes turns RG-irrelevant first, and a patch of zeros subsequently appears in $J_{{\bf k}_N, {\bf k}}$ around this point (Fig.\ref{rgProgression}(left)). Tuning $W/J$ further extends the patch of zeros towards the antinodes (Fig.\ref{rgProgression}(center) and (right)). Kondo screening thus unravels by a systematic decoupling of all $J_{{\bf k}_1, {\bf k}_2}$ that connect adjacent quadrants of the Brillouin zone.   
%{{\textcolor{cyan}{Tuning $W/J$ further extends the patch of zeros in $J_{{\bf k}_1, {\bf k}_2}$ for ${\bf k}_{1}$ lying between a given node and the nearest antinodes (Fig.\ref{rgProgression}(center) and (right)), visualising a systematic decoupling of $J_{{\bf k}_1, {\bf k}_2}$ connecting adjacent quadrants of the Brillouin zone quadrants -- (CHECK THIS SENTENCE, TOO LONG AND SOMEWHAT UNCLEAR)}}.

Precisely at $W/J=(W/J)_{\text{PG}}$, the antinode joins this connected region of zeros in $J_{{\bf k}_1, {\bf k}_2}$, marking the decoupling of the antinodes from all other points in the neighbourhood of the FS. This is an interaction-driven Lifshitz transition of the FS, and marks the entry into a PG phase possessing Fermi arcs~\cite{WuFerrero2018}. Importantly, it coincides with an emergent two-channel Kondo (2CK) impurity model, where each channel corresponds to a pair of Fermi arcs on opposite faces of the conduction bath FS. The 2CK nature of the PG is guaranteed by the symmetry: 
$J_{{\bf k},{\bf k'}}= -J_{{\bf k}+{\bf Q},{\bf k'}}=-J_{{\bf k},{\bf k'}+{\bf Q}}$, where ${\bf Q} = (\pi,\pi)$. The PG expands by shrinking these disconnected Fermi arcs towards the respective nodes, leading to nodal metals whose disappearance heralds the Mott transition.

\par\noindent\textit{{\bf k}-resolved dynamical spectral weight transfer.}
 Strong charge fluctuations develop between the nodal and antinodal regions of the FS in the PG regime (Fig.\ref{chargeCorr}(left)), removing low-energy spectral weight from the antinodes to higher energies and leading to selective gap formation. Accordingly, the PG coincides with the appearance of poles of the lattice model self-energy $\Sigma ({\bf k},\omega=0)$ near the antinodes; these poles approach the nodes on tuning towards the Mott transition (Fig.\ref{chargeCorr}(right)). This mirrors the coalescing of self-energy poles at zero frequency in the underlying impurity model~\cite{suppmat}. \textcolor{blue}{This process corresponds to the formation of Luttinger surfaces of zeros of the single-particle Greens function (CITE Dzyaloshinskii PRB  68, 085113 (2003)) whose growth characterises the PG, and connection to the Mott insulator.} 
\begin{figure}
    \centering
    \includegraphics[width=0.49\linewidth]{cfnode-2.pdf}
    \includegraphics[width=0.49\linewidth]{selfEnergyKspace-3.pdf}
    \caption{Left: Enhanced charge isospin correlations $\chi_c({\bf k}_1, {\bf k}_2) = \braket{c^\dagger_{{\bf k}_1\uparrow}c^\dagger_{{\bf k}_1\downarrow}c_{{\bf k}_2\downarrow}c_{{\bf k}_2\uparrow} + \text{h.c.}}$ between the nodal and antinodal regions, signalling Kondo breakdown in the pseudogap phase of the impurity model. Right: In turn, the breakdown leads to the gapping of the antinodal regions in the lattice model, seen from the appearance of poles in the imaginary part of the self-energy.}
    \label{chargeCorr}
\end{figure}
